---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Inital Setup

You might need to change the working directory before

Be sure to remove all variables from the environment
```{r setup}
rm(list=ls(all=TRUE)) 
```

Now load all necessary libraries
``` {r, results = 'hide'}
suppressWarnings(library(knitr))
suppressMessages(library(data.table))
suppressMessages(library(ggplot2))
```

set knitr options
``` {r}
opts_chunk$set(echo = TRUE, results = 'hold')
```

## Loading and preprocessing the data

``` {r}
# unzip activity file
unzip("activity.zip")

# reading all the data from csv file
acdat = read.csv('activity.csv', header = T)
```

look at the loaded data

``` {r}
names(acdat)
```

``` {r}
str(acdat)
```

``` {r}
head(acdat)
```

## What is mean total number of steps taken per day?

### create a histogram of the total number of steps taken for each day

``` {r}
stepdata <- aggregate(steps ~ date, data=acdat, FUN=sum)
barplot(stepdata$steps, names.arg=stepdata$date, xlab="date", ylab="steps")
```

### Calculate and report the mean and median total number of steps taken per day

``` {r}
mean(stepdata$steps)
median(stepdata$steps)
```


## What is the average daily activity pattern?

Now create a time-series chart. The interval is set to 5-minutes for the x-axis. The avg steps as the y-axis

``` {r}
intervaldata <- aggregate(steps ~ interval, data = acdat, FUN = mean)
plot(intervaldata, type = "l")
```

## Imputing missing values

Fill in the gaps with the mean values of the 5-minute intervals

``` {r}
# Create a new dataset (activity) that is equal to the original dataset (acdat)
activity <- merge(acdat, intervaldata, by="interval", suffixes=c("",".y"))

# find missing data
missingdata <- is.na(activity$steps)

# fill missing date 
activity$steps[missingdata] <- activity$steps.y[missingdata]
activity <- activity[,c(1:3)]
```

Now make a histogram of the total number of steps taken each day. Then calculate and report the mean and median total number of steps taken per day. Does these values differ from the estimates of the first part? 

``` {r}
stepdata <- aggregate(steps ~ date, data=activity, FUN=sum)
barplot(stepdata$steps, names.arg=stepdata$date, xlab="date", ylab="steps")
mean(stepdata$steps)
median(stepdata$steps)
```

## Are there differences in activity patterns between weekdays and weekends?

Create some new variable daytype

``` {r}
# convert to real date type
activity$dateValue <- as.POSIXlt(activity$date,format="%Y-%m-%d")

# set the daytype depending to the day of the week
activity$daytype <- "weekday"
activity$daytype[weekdays(activity$dateValue) %in% c("Saturday","Sunday")] <- "weekend"
``` 

now compare  the weekday and weekend data

``` {r}
data <- aggregate(steps ~ interval, data = activity, subset = activity$daytype == "weekend", FUN = mean)
plot(data, type = "l", main = "weekend")

data <- aggregate(steps ~ interval, data = activity, subset = activity$daytype == "weekday", FUN = mean)
plot(data, type = "l", main = "weekday")
```